Partitioning growth-dependent from growth-independent influences on tropical tree mortality

### 2018-08-23 ms/manuscript.Rnw
Final set of changes

% 250 words
\begin{abstract}
Tree death drives population dynamics, nutrient cycling, and evolution within plant communities. Mortality variation across species is thought to be influenced by different factors relative to variation within species. The unified model provided here separates mortality rates into growth-dependent and growth-independent hazards. This model creates the opportunity to simultaneously estimate the descriptors of variation both across and within species. Moreover, it provides the ability to examine how species traits affect growth-dependent and growth-independent hazards. We derive this unified mortality model  using cross-validated Bayesian methods coupled with mortality data collected over three census intervals for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. We found that growth independent mortality tended to be higher in species with lower wood density, higher light requirements and smaller maximum dbh. Mortality due to marginal carbon budget as measured by near-zero growth rate tended to be higher in species with lower wood density and higher light demand. The total mortality variation attributable to differences among species was large relative to variation explained by these traits, emphasizing that much remains to be understood. This statistical framework strengthens our capacity to parse and understand individual-level mortality in highly diverse tropical forests, and hence to predict its consequences. 
\end{abstract}

@@

Here we propose the following hazard function, incorporating individual-, species-, and census-period effects:
\begin{equation} \label{eq:model}
\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Growth independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{i}(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
\end{equation}
Eq. \ref{eq:model} allows for additive growth-dependent and growth-independent effects, includes a negative exponential effect of the individual's growth rate ($X$; our measure of carbon budget), allows the overall mortality rate to vary among censuses via the random effect $\delta_{t}$, and also provides for species-level differences in the shape of $\lambda_i(t)$ via the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$. The species-level parameters are in turn modelled as a function of both species trait effects and species-level random effects, the latter of which accounts for residual species variation not attributed to traits.

A further advantage of the proposed hazard function (eq. {eq:model}) is that it allows us to estimate (at population level), the relative contributions of different mortality events, even if the causes of death for each individual tree remain unknown.

@@

Effects of species' wood density, maximum size and light demand are now being incorporated into mortality algorithms within many vegetation models in the effort to account for among-species variation \cite{Moorcroft:2001ws,Scheiter:2013ed}. However, the present analysis indicates that such traits are likely to capture only a small proportion of the overall inter-specific variation (Fig. \ref{fig4}, Table~\ref{table:1}). This means that dynamic vegetation models incorporating these traits are likely to underestimate the true variability in tropical rainforest mortality rates, and as a consequence underestimate carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.

>>>>

% 250 words
\begin{abstract}
Tree death drives population dynamics, nutrient cycling, and evolution within plant communities. Mortality variation across species is thought to be influenced by different factors relative to variation within species. The unified model provided here separates mortality rates into growth-dependent and growth-independent hazards. This model creates the opportunity to simultaneously estimate the descriptors of variation both across and within species. Moreover, it provides the ability to examine how species traits affect growth-dependent and growth-independent hazards. We derive this unified mortality model  using cross-validated Bayesian methods coupled with mortality data collected over three census intervals for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama. We found that growth independent mortality tended to be higher in species with lower wood density, higher light requirements and smaller maximum dbh. Mortality due to marginal carbon budget as measured by near-zero growth rate tended to be higher in species with lower wood density and higher light demand. The total mortality variation attributable to differences among species was large relative to variation explained by these traits, emphasizing that much remains to be understood. This additive hazards model strengthens our capacity to parse and understand individual-level mortality in highly diverse tropical forests, and hence to predict its consequences. 
\end{abstract}

@@

Here we propose the following hazard function, incorporating individual-, species-, and census-period effects:
\begin{equation} \label{eq:model}
\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Growth independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{i}(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
\end{equation}

Eq. \ref{eq:model} allows for additive growth-dependent and growth-independent effects, includes a negative exponential effect of the individual's growth rate ($X$; our measure of carbon budget), allows the overall mortality rate to vary among censuses via the random effect $\delta_{t}$, and also provides for species-level differences in the shape of $\lambda_i(t)$ via the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$. The species-level parameters are in turn modelled as a function of both species trait effects and species-level random effects, the latter of which accounts for residual species variation not attributed to traits.

A further advantage of the proposed hazard function (eq. \ref{eq:model}) is that it allows us to estimate (at population level), the relative contributions of different mortality events, even if the causes of death for each individual tree remain unknown.

@@

Basal area growth is thought to be intrinsically linked to changes in total leaf area \cite{Falster2018xx}, and thus, a better predictor of tree mortality. However, we found the opposite -- past diameter growth better predicted tropical tree mortality relative to past basal area growth. This finding may be due to diameter growth being more variable among smaller, more vulnerable, tree sizes (Fig. S8), and thus, better able to capture size-dependent mortality. We did not directly account for individual size in our additive hazards model. This was because of two main reasons: 1) size is highly correlated with both measures of growth (pearson correlations up to 75\%; Fig S8), and consequently is prone to parameter identifiability issues where the model is unable to discern growth from size effects; 2) the effect of size is highly non-linear \cite{Vieilledent:2010fv,Ruger:2011cv}, and thus, its inclusion and subsequent interpretation would not be straight forward. Further work is required in order to determine the independent effects of size and growth effects on tree mortality, but also how to partition sizes effects between growth-dependent and growth-independent hazards.

Effects of species' wood density, maximum size and light demand are now being incorporated into mortality algorithms within many vegetation models in the effort to account for among-species variation \cite{Moorcroft:2001ws,Scheiter:2013ed}. However, the present analysis indicates that such traits are likely to capture only a small proportion of the overall inter-specific variation (Fig. \ref{fig4}, Table~\ref{table:1}). This means that dynamic vegetation models incorporating these traits are likely to account for a fraction of the true variability in tropical rainforest mortality rates, and as a consequence underestimate carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.
